---
title: "Single cell mouse RNA-seq classification"
output:
  html_document:
    df_print: paged
---
Load libraries
```{r}
library(caret, suppressWarnings(TRUE))
```

Load data
```{r}
mouse_data <- read.table("mouse_data.csv", sep = ",", header = TRUE)
```

Delete first column which contain ids
```{r}
mouse_data <- mouse_data[,-1]
```

Explore data set
```{r}
dim(mouse_data)
```

As you can see, it's quite a large data set. There are 3,005 samples each with 19,976 variables. This means the dataframe has 60,027,880 cells! A set this big will not only slow down computation, but it may also cause a memory overflow. Performing a dimentionality reduction (such as PCA) may solve this problem.

Set random seed to achieve reproducible results
```{r}
set.seed(333)
```

Split into train and test sets
```{r}
inTraining <- createDataPartition(mouse_data$level1class, p=.75, list = FALSE)
df_train <- mouse_data[ inTraining,]
df_test <- mouse_data[-inTraining,]
```

Pre-process data
```{r}
# the target variable is level1class in column #3, so we don't want to modify that variable
ppv <- preProcess(df_train[,-3], method = c("center", "scale", "pca", "nzv"))
transformed_train <- predict(ppv, newdata = df_train)
transformed_test <- predict(ppv, newdata = df_test)
```

Let's compare the original training set with the transformed one.
```{r}
dim(df_train)
format(object.size(df_train), units = "Mb")
dim(transformed_train)
format(object.size(transformed_train), units = "Mb")
```

Training the KNN model with cross validation and 5 resampling iterations.
```{r}
knn_fit <- train(level1class ~., data = transformed_train, method = "knn",
                 trControl=trainControl(method = "cv", number = 5),
                 tuneLength = 5)
```

Print the training results
```{r}
knn_fit
```

The highest accuracy was achieved when $k=5$. We can also plot the variation in accuracy with respect to $k$.
```{r}
plot(knn_fit)
```

Now let's validate the model using the test set. First, we'll use the model we constructed in the previous step to make the classification predictions in our training set.
```{r}
test_pred <- predict(knn_fit, newdata = transformed_test)
```

Finally, we compare our predictions with the original labels.

```{r}
confusionMatrix(test_pred, transformed_test$level1class)
```

We mentioned earlier that a data set large data set might cause problems in computation. In this particular case, if we try training the model with the unreduced training set, there is a memory overflow.
```{r}
knn_fit <- train(level1class ~., data = df_train, method = "knn",
                 trControl=trainControl(method = "cv", number = 5),
                 preProcess = c("center", "scale", "nzv"),
                 tuneLength = 5)
```

